---
title: "Assignment 2"
output:
  html_document:
    toc: yes
editor_options:
  chunk_output_type: inline
---  

```{r load packages, warning = FALSE, message = FALSE, echo = FALSE}
# load your packages here:
library(gapminder)
library(tidyverse)
library(knitr)
library(kableExtra)
library(DT)
```

# Exercise 1
## 1.1

```{r}
ex1 <- gapminder %>%
  filter(country=="Bangladesh" | country=="Brazil" | country=="Philippines") %>% 
  filter(year >= 1970 & year <= 1979)
print(ex1)
```

## 1.2

```{r}
ex1 %>% 
  select(country, gdpPercap)
```

## 1.3
```{r}
gapminder_modified <- gapminder %>%
  mutate(lagLifeExp = lag(lifeExp, 11)) %>% 
  mutate(diffLifeExp = lifeExp - lagLifeExp) %>% 
  filter(year==2007) %>% 
  mutate(yn = ifelse(diffLifeExp > 0, "Increase", 
                     ifelse(diffLifeExp < 0, "Decrease", "Constant"))) %>% 
  rename(lifeExp2007 = lifeExp, lifeExp1952 = lagLifeExp) %>% 
  select(country, continent, lifeExp2007, lifeExp1952, diffLifeExp, yn) %>% 
  filter(yn=="Decrease") 
print(gapminder_modified)
```


## 1.4
```{r}
gapminder %>% 
  group_by(country) %>%  
  summarise(maxGDP = max(gdpPercap))
```


## 1.5
```{r, fig.width = 5, fig.height = 2}
gapminder %>% 
  filter(country=="Canada") 
  ggplot(gapminder, aes(gdpPercap, lifeExp)) + 
  geom_point(alpha = 0.1) +
  scale_x_log10("GDP per capita (in log)", labels = scales::dollar_format()) +
  theme_bw() +
  ylab("Life Expectancy")
```


# Exercise 2
```{r}
gapminder_n <- gapminder %>% 
  select(continent, lifeExp)
print(gapminder_n)
```

**Possible values for categorical variable (_continent_)**: Name of continents.

**Possible values for continuous variable (_lifeExp_)**: 0 to infinity (i.e., we can calculate mean, median, minimum, maximum, range, variance, standard deviation, etc.)

## Summary of a categorical variable (_continent_)
### Frequency table
```{r}
gapminder_n1 <- gapminder_n %>% 
  select(continent) %>%
  group_by(continent) %>% 
  summarise (n = n()) %>%
  mutate(freq = n / sum(n))
print(gapminder_n1)
```

### Barplot
```{r}
gapminder_n2 <- gapminder_n1 %>% 
  select(continent, freq) 
  ggplot(gapminder_n1, aes(x = continent, y = freq)) +
    geom_bar(stat = "identity", color = 5, fill = grey(0.7)) + 
    xlab("Continent") +
    ylab("Proportion") +
    geom_text(aes(label=round(freq, 2)), vjust = -0.6, size = 4)+ 
    theme_minimal()
```

### Pie chart
```{r}
ggplot(gapminder_n2, aes(x = "", y = freq, fill = continent)) +
  geom_bar(width = 1, stat = "identity", color = "blue") +
  coord_polar("y", start = 0)+
  geom_text(aes(y = freq/5 + c(0, cumsum(freq)[-length(freq)]), 
            label = round(freq, 2)), size=4) +
  scale_fill_brewer(palette="Blues") +
  theme_void() 
  
```


## Summary of a continuous variable (_lifeExp_)
```{r}
gapminder_nc <- gapminder_n %>%
  select(lifeExp)
```

### Summary statistics
```{r}
gapminder_nc %>% 
  summarise(mean = mean(lifeExp), median = median(lifeExp), variance = var(lifeExp), 
            minimum = min(lifeExp), maximum = max(lifeExp))
```

### Histogram
```{r}
hist(gapminder_nc$lifeExp, xlab = "Life expectancy", freq = F, main = "", ylim = c(0, 0.04),
     xlim = c(20, 90))
```

## Summary of a continuous variable (_lifeExp_) in terms of a categorical variable (_continent_)
### Summary statistics
```{r}
gapminder_n %>% 
  group_by(continent) %>% 
  summarise(mean = mean(lifeExp), median = median(lifeExp), variance = var(lifeExp), 
            minimum = min(lifeExp), maximum = max(lifeExp))
```

### Boxplot
```{r}
ggplot(gapminder_n, aes( x = continent, y = lifeExp)) +
  geom_boxplot() + xlab("Continent") + ylab("Life expectancy")
```

# Exercise 3
## Scatter plot
```{r, fig.width = 5, fig.height = 2}
data(women)
ggplot(women, aes(height, weight)) + 
  geom_point(alpha = 0.1) +
  xlab("Height in inch") +
  ylab("Weight in lb") + 
  theme_bw() 
```

## Scatter plot plus another plot

```{r, fig.width = 5, fig.height = 2, warning = FALSE, message = FALSE}
library(cowplot)
plot1 <- ggplot(women, aes(height, weight)) + 
  geom_point(alpha = 0.1) +
  xlab("Height in inch") +
  ylab("Weight in lb") + 
  theme_bw() 
plot2 <- ggplot(women, aes(height, weight)) + 
  geom_point(alpha = 0.1) +
  xlab("Height in inch") +
  ylab("Weight in lb") + 
  theme_bw() + 
  geom_smooth(method = 'lm', formula = y~x)
plot_grid(plot1, plot2, labels = "AUTO")
```

# Recycling
## Wrong way
```{r}
filter(gapminder, country == c("Rwanda", "Afghanistan"))
```

**Comment**: The analyst wasn't succeeded. The analyst used concatenate with the symbol ',' (comma), but it should be '|' (or).

## Right way
```{r}
filter(gapminder, country=="Rwanda" | country=="Afghanistan")
```

# Tibble display
## knitr::kable()
```{r, warning = FALSE, message = FALSE}
filter(gapminder, country=="Rwanda" & year < 2000) %>% 
knitr::kable() %>%
  kable_styling()
```

## DT::datatable()
```{r, warning = FALSE, message = FALSE}
filter(gapminder, country=="Rwanda" | country=="Afghanistan") %>% 
DT::datatable(gapminder)
```


